---
title: "Project1"
output:
  pdf_document: default
  html_document: default
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R markdown project

This project analyses data from Wooldridge Source: M. Blackburn and D. Neumark (1992), “Unobserved Ability, Efficiency Wages, and Interindustry Wage Differentials,” Quarterly Journal of Economics 107, 1421-1436.

In this exploration, we aim to unpack the hypothesis suggesting that married women command higher wages. As we navigate the intricacies of a linear regression model, our focus extends beyond scrutinizing the "married" variable alone. We delve into not only the coefficient and significance of marital status but also other pertinent correlations with wage, presenting a holistic perspective on the myriad factors influencing earnings.

A data.frame with 935 observations on 17 variables:

wage: monthly earnings

hours: average weekly hours

IQ: IQ score

KWW: knowledge of world work score

educ: years of education

exper: years of work experience

tenure: years with current employer

age: age in years

married: =1 if married

black: =1 if black

south: =1 if live in south

urban: =1 if live in SMSA (Standard Metropolitan Statistical Area)

sibs: number of siblings

brthord: birth order

meduc: mother's education

feduc: father's education

lwage: natural log of wage


```{r wage2, message=FALSE}
library(wooldridge)
library(Hmisc)

data('wage2')
# Removing missing values, total 663 observations
wage2=na.omit(wage2)
#describe(wage2)
head(wage2)
```

## Data visualisation
Creating a scatter matrix plot of data with correlation coefficients.
The scatter plot matrix did not highlight best correlations, that is why providing a separate plot for correlations.


```{r pressure, echo=FALSE, fig.width=12, fig.height=12, include=TRUE, message=FALSE}
library(ggstatsplot)
library(ggplot2)
library(psych)
library(car)
library(lmtest)
library(sandwich)

ggplot(wage2, aes(x = wage2$IQ, y = wage2$wage, color = factor(wage2$married))) +
  geom_point() +
  labs(title = "Scatter Plot of IQ and Wage Colored By Married Status",
       x = "IQ",
       y = "Wage",
       color = "Married Status") +
  scale_color_manual(values = c("green", "blue"))

ggplot(wage2, aes(x = wage2$IQ, y = wage2$wage, color = factor(wage2$black))) +
  geom_point() +
  labs(title = "Scatter Plot of IQ and Wage Colored By Race",
       x = "IQ",
       y = "Wage",
       color = "Race") +
  scale_color_manual(values = c("red", "black"))


ggplot(wage2, aes(x = wage2$IQ, y = wage2$wage, color = factor(wage2$urban))) +
  geom_point() +
  labs(title = "Scatter Plot of IQ and Wage Colored By Urban",
       x = "IQ",
       y = "Wage",
       color = "Urban") +
  scale_color_manual(values = c("blue", "orange"))

# scatter plot matrix with correlations
pairs.panels(wage2, main = "Scatter Plot Matrix for Wage2 Dataset", pch=".", lm=TRUE, smoother=TRUE)

# correlogram
ggstatsplot::ggcorrmat(
  data = wage2,
  type = "parametric", # parametric for Pearson, nonparametric for Spearman's correlation
  colors = c("darkred", "white", "steelblue") # change default colors
)
```

## Normality
Checking the normality of the data.
This block of code checks for normality of the data. Wage is skewed to left as expected, but the log of wage is more towards normal distribution

```{r Normality, message=FALSE}
library('ggpubr')
par(mfrow=c(2,2))
shapiro.test(wage2$wage)
shapiro.test(wage2$lwage)

plot0 <- ggdensity(wage2$wage, main="Density plot of wage", xlab="wage")
plot1 <- ggqqplot(wage2$wage, title="QQ plot of wage")

plot2 <- ggdensity(wage2$lwage, main="Density plot of log wage", xlab="log(wage)")
plot3 <- ggqqplot(wage2$lwage, title="QQ plot of log wage")

ggarrange(plot0, plot1, plot2, plot3, ncol=2, nrow=2)
```
## Linear regression
```{r ,message = FALSE}

# Remove the wage from the model
wage2 <- subset(wage2, select = -wage)

# Standard model
model0 <- lm(wage2$lwage~., data = wage2)

library(MASS)
# step select smaller model
fit <-  lm(wage2$lwage~.,data = wage2)
model1 <- stepAIC(fit, direction='both', trace=0)

summary(model0)
summary(model1)
```

### Collinearity
if higher than 5 indicates collinearity. Here we do not have this issue, but if we did, we would remove one of the variables with high VIF from the model.
```{r Collinearity}
library(car)
vif(model0)
vif(model1)
```

### Heteroscedasticity
```{r Heteroscedasticity}
bptest(model1)
model1.fixed <- coeftest(model1, vcov. = vcovHC(model1, type = 'HC3'))
model1.fixed


bptest(model0)
model0.fixed <- model1.fixed <- coeftest(model1, vcov. = vcovHC(model1, type = 'HC3'))
model0.fixed
```
### Normality of residuals
```{r Normality of residuals}
library(olsrr)

ols_test_normality(model0)
ols_test_normality(model1)

```

### Conclusion
The coefficient for "married" stands at 0.0357696, suggesting a positive association between marital status and wages. Crucially, the p-value of 0.023591, less than the conventional significance level of 0.05, signifies statistical significance. This implies that, according to the model, being married is not merely incidental but significantly associated with higher earnings. Beyond the exploration of marital status, other variables like wage levels, working hours, tenure, and regional dynamics emerge as influential contributors to the earnings tapestry.

In conclusion, the analysis lends support to the hypothesis that married women do, indeed, earn higher wages. The positive coefficient and statistically significant p-value imply a meaningful connection between marital status and increased earnings, providing empirical evidence to corroborate the initial claim. It's imperative to note that this statistical evidence, while robust, does not establish causation, and other unconsidered factors may contribute to the observed relationship.